The Proofs

### 2024-07-16 proofs/tex/alg-pos-subtyping-lemmas.tex
WIP: fix references, add appendix

            \begin{itemize}
                \item $[[nf(uN)]]$ and $[[nf(iM)]]$ are normalized 
                \item $[[Γ ; dom(Θ) ⊢  nf(uN)]]$ because $[[Γ ; dom(Θ) ⊢  uN]]$ (\cref{corollary:wf-nf-algo})
                \item $[[Γ ⊢ nf(iM)]]$ because $[[Γ ⊢ iM]]$ (\cref{corollary:wf-nf})
                \item $[[ Θ ⊢ nf(uσ) : uv(uP) ]]$ because $[[Θ ⊢ uσ : uv(uP) ]]$ (\cref{corollary:norm-subst-sig-algo})
                \item $ \begin{aligned}[t]
                        [[ Γ ⊢ [uσ]uN ≈ iM ]] &\Rightarrow [[ [uσ]uN ≈ iM ]]
                                            && \text {by \cref{lemma:equiv-completeness}}\\
                                            &\Rightarrow [[ nf([uσ]uN) = nf(iM) ]]
                                            && \text {by \cref{lemma:normalization-completeness}}\\
                                            &\Rightarrow [[ [nf(uσ)]nf(uN) = nf(iM) ]]
                                            && \text {by \cref{lemma:norm-subst-distr}}\\
                        \end{aligned}
                    $
            \end{itemize}
            By the completeness of the unification,
            $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ exists, and
            $[[Θ ⊢ nf(uσ) : lift UC]]$,
            and by \cref{corollary:sat-equiv}, $[[ Θ ⊢ uσ : lift UC ]]$.

>>>>

            \begin{itemize}
                \item $[[nf(uN)]]$ and $[[nf(iM)]]$ are normalized 
                \item $[[Γ ; dom(Θ) ⊢  nf(uN)]]$ because $[[Γ ; dom(Θ) ⊢  uN]]$ (\cref{corollary:wf-nf-algo})
                \item $[[Γ ⊢ nf(iM)]]$ because $[[Γ ⊢ iM]]$ (\cref{corollary:wf-nf})
                \item $[[ Θ ⊢ nf(uσ) : uv(uP) ]]$ because $[[Θ ⊢ uσ : uv(uP) ]]$ (\cref{corollary:norm-subst-sig-algo})
                \item \begin{align*}
                        [[ Γ ⊢ [uσ]uN ≈ iM ]] &\Rightarrow [[ [uσ]uN ≈ iM ]]
                                            && \text {by \cref{lemma:equiv-completeness}}\\
                                            &\Rightarrow [[ nf([uσ]uN) = nf(iM) ]]
                                            && \text {by \cref{lemma:normalization-completeness}}\\
                                            &\Rightarrow [[ [nf(uσ)]nf(uN) = nf(iM) ]]
                                            && \text {by \cref{lemma:norm-subst-distr}}\\
                        \end{align*}
            \end{itemize}
            By the completeness of the unification,
            $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ exists, and
            $[[Θ ⊢ nf(uσ) : lift UC]]$,
            and by \cref{corollary:sat-equiv}, $[[ Θ ⊢ uσ : lift UC ]]$.

### 2024-07-16 proofs/tex/decl-subtyping-lemmas.tex
WIP: fix references, add appendix

\lemmaFvPropagation*
\begin{proof}
  Mutual induction on $[[Γ ⊢ iN ≤ iM]]$ and $[[Γ ⊢ iP ≥ iQ]]$.
  \begin{caseof}
  \item $[[G ⊢ a⁻ ≤ a⁻]]$\\
    It is self-evident that $[[{a⁻} ⊆ {a⁻}]]$.
  \item $[[G ⊢ ↑iP ≤ ↑iQ]]$
    From the inversion (and unfolding $[[G ⊢ iP ≈ iQ]]$ ), we have
    $[[G ⊢ iP ≥ iQ]]$. Then by the induction hypothesis,
    $[[fv(iP)]] \subseteq [[fv(iQ)]]$. The desired 
    inclusion holds, since $[[fv(↑iP)]] = [[fv(iP)]]$ and
    $[[fv(↑iQ)]] = [[fv(iQ)]]$.
  \item $[[G ⊢ iP → iN ≤ iQ → iM]]$
    The induction hypothesis applied to the premises gives:
    $[[fv(iP)]] \subseteq [[fv(iQ)]]$ and
    $[[fv(iN)]] \subseteq [[fv(iM)]]$.
    Then $[[fv(iP → iN)]] = [[fv(iP) ∪ fv(iN)]] \subseteq
    [[fv(iQ) ∪ fv(iM)]] = [[fv(iQ → iM)]]$.

  \item $[[G ⊢ ∀pas.iN ≤ ∀pbs.iM]]$\\
    $
    \begin{aligned}[t]
      [[fv ∀pas.iN ]] &\subseteq [[fv ([iPs/pas] iN) ]] ~\setminus~ [[{pbs}]] 
                      &&   \text{here $[[{pbs}]]$ is excluded by the premise $[[fv iN ∩ {pbs} = ∅]]$}\\
                      &\subseteq [[fv iM]] ~\setminus~ [[{pbs}]]
                      &&   \text{by the induction hypothesis, } [[fv ([iPs/pas] iN) ]] \subseteq [[fv iM]] \\
                      &\subseteq [[fv ∀pbs.iM]]
    \end{aligned}
    $
  \item The positive cases are symmetric.
  \end{caseof}
\end{proof}

@@

      Notice that we can weaken and rearrange the contexts without changing the sizes of the 
      derivations: $[[Γ, {pbs}, {pcs} ⊢ [iPs/pas]iN' ≤ iM1']]$
      and $[[Γ, {pbs}, {pcs} ⊢ [iQs/pbs]iM2' ≤ iK']]$. This way, 
      the sum of the sizes of these derivations is smaller than the sum of the sizes of
      $[[Γ ⊢ iN ≤ iM1]]$ and $[[Γ ⊢ iM2 ≤ iK]]$.
      Let us apply the induction hypothesis to these derivations, 
      with the substitution $[[ Γ', pcs ⊢ σ ○ (iQs/pbs) : Γ, {pbs}, {pcs}  ]]$
      (\cref{lemma:subst-domain-weakening}).
      To apply the induction hypothesis, it is left to show that 
      $[[ σ ○ (iQs/pbs) ]]$ unifies $[[iM1']]$ and $[[ [iQs/pbs]iM2']]$:
      $$
      \begin{aligned}[t]
        [[ [σ ○ iQs/pbs]iM1' ]] &= [[ [σ][iQs/pbs]iM1' ]]\\
                                &= [[ [ [σ]iQs/pbs ][σ]iM2' ]]
                                && \text{by \cref{lemma:subst-composition}}\\
                                &= [[ [ [σ]iQs/pbs ][σ]iM2' ]]
                                && \text{Since $[[ [σ]iM1' = [σ]iM2' ]]$}\\
                                &= [[  [σ][iQs/pbs]iM2' ]]
                                && \text{by \cref{lemma:subst-composition}}\\
                                &= [[  [σ][iQs/pbs][iQs/pbs]iM2' ]]
                                && \text{Since $[[Γ, pcs ⊢ iQs]]$, and $[[{(Γ, pcs)} ∩ {pbs} = ∅]]$ }\\
                                &= [[  [σ ○ iQs/pbs][iQs/pbs]iM2' ]]
      \end{aligned}
      $$
      This way the induction hypothesis gives us
      $[[ Γ', pcs ⊢ [σ][iQs/pbs][iPs/pas]iN' ≤  [σ][iQs/pbs]iK' ]]$,
      and since $[[Γ, pcs ⊢ iK']]$, $[[ [iQs/pbs]iK' = iK' ]]$, that is 
      $[[ Γ', pcs ⊢ [σ][iQs/pbs][iPs/pas]iN' ≤  [σ]iK' ]]$.
      Let us rewrite the substitution that we apply to $[[iN']]$:
      $$
      \begin{aligned}[t]
        [[ [σ ○ iQs/pbs ○ iPs/pas]iN' ]] &= [[ [ (σ <=< iQs/pbs) ○ σ ○ iPs/pas]iN' ]]
                                       && \text{by \cref{lemma:subst-composition}}\\
                                       &= [[ [(σ <=< iQs/pbs) ○ (σ <=< iPs/pas) ○ σ] iN' ]]
                                       && \text{by \cref{lemma:subst-composition}}\\
                                       &= [[ [(((σ <=< iQs/pbs) ○ σ) <=< iPs/pas) ○ σ] iN' ]]
                                       && \text{Since $[[fv([σ]iN') ∩ {pbs} = ∅]]$}\\
                                       &= [[ [((σ ○ iQs/pbs) <=< iPs/pas) ○ σ] iN' ]]
                                       && \text{by \cref{lemma:subst-composition}}\\
                                       &= [[ [(σ ○ iQs/pbs) <=< iPs/pas][σ] iN' ]]
      \end{aligned}
      $$
      Notice that $[[(σ ○ iQs/pbs) <=< iPs/pas]]$
      is a substitution that turns $[[pai]]$ into $[[ [σ ○ iQs/pbs]iPi ]]$, 
      where $[[ Γ',pcs ⊢ [σ ○ iQs/pbs]iPi]]$.
      This way, 
      $[[ Γ', pcs ⊢ [(σ ○ iQs/pbs) <=< iPs/pas][σ]iN' ≤  [σ]iK' ]]$
      means $[[Γ ⊢ ∀pas.[σ]iN' ≤ ∀pcs.[σ]iK']]$
      by \cref{lemma:quant-rule-decomposition}, that is
      $[[Γ ⊢ [σ]iN ≤ [σ]iK]]$, as required.
  \end{caseof}

>>>>

\lemmaFvPropagation*
\begin{proof}
  Mutual induction on $[[Γ ⊢ iN ≤ iM]]$ and $[[Γ ⊢ iP ≥ iQ]]$.
  \begin{caseof}
  \item $[[G ⊢ a⁻ ≤ a⁻]]$\\
    It is self-evident that $[[{a⁻} ⊆ {a⁻}]]$.
  \item $[[G ⊢ ↑iP ≤ ↑iQ]]$
    From the inversion (and unfolding $[[G ⊢ iP ≈ iQ]]$ ), we have
    $[[G ⊢ iP ≥ iQ]]$. Then by the induction hypothesis,
    $[[fv(iP)]] \subseteq [[fv(iQ)]]$. The desired 
    inclusion holds, since $[[fv(↑iP)]] = [[fv(iP)]]$ and
    $[[fv(↑iQ)]] = [[fv(iQ)]]$.
  \item $[[G ⊢ iP → iN ≤ iQ → iM]]$
    The induction hypothesis applied to the premises gives:
    $[[fv(iP)]] \subseteq [[fv(iQ)]]$ and
    $[[fv(iN)]] \subseteq [[fv(iM)]]$.
    Then $[[fv(iP → iN)]] = [[fv(iP) ∪ fv(iN)]] \subseteq
    [[fv(iQ) ∪ fv(iM)]] = [[fv(iQ → iM)]]$.
  \item $[[G ⊢ ∀pas.iN ≤ ∀pbs.iM]]$
    \begin{align*}
      [[fv ∀pas.iN ]] &\subseteq [[fv ([iPs/pas] iN) ]] ~\setminus~ [[{pbs}]] 
                      &&   \text{$[[{pbs}]]$ is excluded by the premise $[[fv iN ∩ {pbs} = ∅]]$}\\
                      &\subseteq [[fv iM]] ~\setminus~ [[{pbs}]]
                      &&   \text{by the i.h., } [[fv ([iPs/pas] iN) ]] \subseteq [[fv iM]] \\
                      &\subseteq [[fv ∀pbs.iM]]
    \end{align*}
  \item The positive cases are symmetric.
  \end{caseof}
\end{proof}

@@

      Notice that we can weaken and rearrange the contexts without changing the sizes of the 
      derivations: $[[Γ, {pbs}, {pcs} ⊢ [iPs/pas]iN' ≤ iM1']]$
      and $[[Γ, {pbs}, {pcs} ⊢ [iQs/pbs]iM2' ≤ iK']]$. This way, 
      the sum of the sizes of these derivations is smaller than the sum of the sizes of
      $[[Γ ⊢ iN ≤ iM1]]$ and $[[Γ ⊢ iM2 ≤ iK]]$.
      Let us apply the induction hypothesis to these derivations, 
      with the substitution $[[ Γ', pcs ⊢ σ ○ (iQs/pbs) : Γ, {pbs}, {pcs}  ]]$
      (\cref{lemma:subst-domain-weakening}).
      To apply the induction hypothesis, it is left to show that 
      $[[ σ ○ (iQs/pbs) ]]$ unifies $[[iM1']]$ and $[[ [iQs/pbs]iM2']]$:

      \begin{align*}
        [[ [σ ○ iQs/pbs]iM1' ]] &= [[ [σ][iQs/pbs]iM1' ]]\\
                                &= [[ [ [σ]iQs/pbs ][σ]iM2' ]]
                                && \text{by \cref{lemma:subst-composition}}\\
                                &= [[ [ [σ]iQs/pbs ][σ]iM2' ]]
                                && \text{Since $[[ [σ]iM1' = [σ]iM2' ]]$}\\
                                &= [[  [σ][iQs/pbs]iM2' ]]
                                && \text{by \cref{lemma:subst-composition}}\\
                                &= [[  [σ][iQs/pbs][iQs/pbs]iM2' ]]
                                && \text{Since $[[Γ, pcs ⊢ iQs]]$, and $[[{(Γ, pcs)} ∩ {pbs} = ∅]]$ }\\
                                &= [[  [σ ○ iQs/pbs][iQs/pbs]iM2' ]]
      \end{align*}

      This way the induction hypothesis gives us
      $[[ Γ', pcs ⊢ [σ][iQs/pbs][iPs/pas]iN' ≤  [σ][iQs/pbs]iK' ]]$,
      and since $[[Γ, pcs ⊢ iK']]$, $[[ [iQs/pbs]iK' = iK' ]]$, that is 
      $[[ Γ', pcs ⊢ [σ][iQs/pbs][iPs/pas]iN' ≤  [σ]iK' ]]$.
      Let us rewrite the substitution that we apply to $[[iN']]$:
      \begin{align*}
        [[ [σ ○ iQs/pbs ○ iPs/pas]iN' ]] &= [[ [ (σ <=< iQs/pbs) ○ σ ○ iPs/pas]iN' ]]
                                       && \text{by \cref{lemma:subst-composition}}\\
                                       &= [[ [(σ <=< iQs/pbs) ○ (σ <=< iPs/pas) ○ σ] iN' ]]
                                       && \text{by \cref{lemma:subst-composition}}\\
                                       &= [[ [(((σ <=< iQs/pbs) ○ σ) <=< iPs/pas) ○ σ] iN' ]]
                                       && [[fv([σ]iN') ∩ {pbs} = ∅]]\\
                                       &= [[ [((σ ○ iQs/pbs) <=< iPs/pas) ○ σ] iN' ]]
                                       && \text{by \cref{lemma:subst-composition}}\\
                                       &= [[ [(σ ○ iQs/pbs) <=< iPs/pas][σ] iN' ]]
      \end{align*}
      Notice that $[[(σ ○ iQs/pbs) <=< iPs/pas]]$
      is a substitution that turns $[[pai]]$ into $[[ [σ ○ iQs/pbs]iPi ]]$, 
      where $[[ Γ',pcs ⊢ [σ ○ iQs/pbs]iPi]]$.
      This way, 
      $[[ Γ', pcs ⊢ [(σ ○ iQs/pbs) <=< iPs/pas][σ]iN' ≤  [σ]iK' ]]$
      means $[[Γ ⊢ ∀pas.[σ]iN' ≤ ∀pcs.[σ]iK']]$
      by \cref{lemma:quant-rule-decomposition}, that is
      $[[Γ ⊢ [σ]iN ≤ [σ]iK]]$, as required.
  \end{caseof}

### 2024-07-16 proofs/tex/equiv-lemmas.tex
WIP: fix references, add appendix

\lemmaWfEquiv*
\begin{proof}
  We prove it by induction on $[[iP ≈ iQ]]$ and mutually, on $[[iN ≈ iM]]$.
  Let us consider the last rule used in the derivation.
  \begin{caseof}
    \item \ruleref{\ottdruleEOneNVarLabel}, 
      that is $[[iN ≈ iM]]$ has shape $[[α⁻ ≈ α⁻]]$.\\
      Than $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ is rewritten as $[[Γ ⊢ α⁻]] \iff [[Γ ⊢ α⁻]]$,
      which holds trivially.
    \item \ruleref{\ottdruleEOneShiftULabel},
      that is $[[iN ≈ iM]]$ has shape $[[↑iP ≈ ↑iQ]]$.\\
      By inversion, $[[iP ≈ iQ]]$, and by the induction hypothesis,
      $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$. 
      Also notice that $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]]$ 
      and $[[Γ ⊢ ↑iQ]] \iff [[Γ ⊢ iQ]]$ by inversion and \ruleref{\ottdruleWFTShiftULabel}.
      This way, $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]] \iff [[Γ ⊢ ↑iQ]]$.
    \item \ruleref{\ottdruleEOneArrowLabel},
      that is $[[iN ≈ iM]]$ has shape $[[iP → iN' ≈ iQ → iM']]$.\\
      Then by inversion, $[[iP ≈ iQ]]$ and $[[iN' ≈ iM']]$, 
      and by the induction hypothesis, $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ and $[[Γ ⊢ iN']] \iff [[Γ ⊢ iM']]$.

      $
      \begin{aligned}[t]
        [[Γ ⊢ iP → iN']]  &\iff [[Γ ⊢ iP]] \text{ and } [[Γ ⊢ iN']] 
                          && \text{by inversion and \ruleref{\ottdruleWFTArrowLabel}}\\
                          &\iff [[Γ ⊢ iQ]] \text{ and } [[Γ ⊢ iM']]
                          && \text{as noted above}\\
                          &\iff [[Γ ⊢ iQ → iM']]
                          && \text{by \ruleref{\ottdruleWFTArrowLabel} and inversion}\\
      \end{aligned}
      $\\
    \item \ruleref{\ottdruleEOneForallLabel},
      that is $[[iN ≈ iM]]$ has shape $[[∀pas.iN' ≈ ∀pbs.iM']]$.\\
      By inversion, $[[∀pas.iN' ≈ ∀pbs.iM']]$ means that
      $[[{pas} ∩ fv iM = ∅]]$ and 
      that there exists a bijection on variables 
      $[[mu : ({pbs} ∩ fv iM') ↔ ({pas} ∩ fv iN')]]$
      such that $[[iN' ≈ [mu] iM']]$.

@@

    To show that $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$, let us take an arbitrary
    $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$.

    $
    \begin{aligned}[t]
      [[β±]] &= [[ [id]β± ]]
             && \text{by definition of $[[id]]$}\\
             &\eqDOne [[ [σ1']β± ]]
             && \text{by \ref{fact:subs-proper-sub:forall-ih}}\\
             &= [[ [iQs/pas][σ1]β±]]
             && \text{by definition of $[[σ1']]$}\\
             &= [[ [σ1]β± ]]
             && \text{because $[[{pas} ∩ fv [σ1]β±]] \subseteq [[{pas} ∩ Γ]] = \emptyset$}
    \end{aligned}
    $\\
    This way, $[[Γ ⊢ [σ1]β± ≈ β±]]$ for any $[[β±]] \in [[fv ∀pas.iM]]$ and thus,
    $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$.

>>>>

\lemmaWfEquiv*
\begin{proof}
  We prove it by induction on $[[iP ≈ iQ]]$ and mutually, on $[[iN ≈ iM]]$.
  Let us consider the last rule used in the derivation.
  \begin{caseof}
    \item \ruleref{\ottdruleEOneNVarLabel}, 
      that is $[[iN ≈ iM]]$ has shape $[[α⁻ ≈ α⁻]]$.\\
      Than $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ is rewritten as $[[Γ ⊢ α⁻]] \iff [[Γ ⊢ α⁻]]$,
      which holds trivially.
    \item \ruleref{\ottdruleEOneShiftULabel},
      that is $[[iN ≈ iM]]$ has shape $[[↑iP ≈ ↑iQ]]$.\\
      By inversion, $[[iP ≈ iQ]]$, and by the induction hypothesis,
      $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$. 
      Also notice that $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]]$ 
      and $[[Γ ⊢ ↑iQ]] \iff [[Γ ⊢ iQ]]$ by inversion and \ruleref{\ottdruleWFTShiftULabel}.
      This way, $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]] \iff [[Γ ⊢ ↑iQ]]$.
    \item \ruleref{\ottdruleEOneArrowLabel},
      that is $[[iN ≈ iM]]$ has shape $[[iP → iN' ≈ iQ → iM']]$.\\
      Then by inversion, $[[iP ≈ iQ]]$ and $[[iN' ≈ iM']]$, 
      and by the induction hypothesis, $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ and $[[Γ ⊢ iN']] \iff [[Γ ⊢ iM']]$.
      \begin{align*}
        [[Γ ⊢ iP → iN']]  &\iff [[Γ ⊢ iP]] \text{ and } [[Γ ⊢ iN']] 
                          && \text{by inversion and \ruleref{\ottdruleWFTArrowLabel}}\\
                          &\iff [[Γ ⊢ iQ]] \text{ and } [[Γ ⊢ iM']]
                          && \text{as noted above}\\
                          &\iff [[Γ ⊢ iQ → iM']]
                          && \text{by \ruleref{\ottdruleWFTArrowLabel} and inversion}\\
      \end{align*}
    \item \ruleref{\ottdruleEOneForallLabel},
      that is $[[iN ≈ iM]]$ has shape $[[∀pas.iN' ≈ ∀pbs.iM']]$.\\
      By inversion, $[[∀pas.iN' ≈ ∀pbs.iM']]$ means that
      $[[{pas} ∩ fv iM = ∅]]$ and 
      that there exists a bijection on variables 
      $[[mu : ({pbs} ∩ fv iM') ↔ ({pas} ∩ fv iN')]]$
      such that $[[iN' ≈ [mu] iM']]$.

@@

    To show that $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$, let us take an arbitrary
    $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$.
    \begin{align*}
      [[β±]] &= [[ [id]β± ]]
             && \text{by definition of $[[id]]$}\\
             &\eqDOne [[ [σ1']β± ]]
             && \text{by \ref{fact:subs-proper-sub:forall-ih}}\\
             &= [[ [iQs/pas][σ1]β±]]
             && \text{by definition of $[[σ1']]$}\\
             &= [[ [σ1]β± ]]
             && \text{because $[[{pas} ∩ fv [σ1]β±]] \subseteq [[{pas} ∩ Γ]] = \emptyset$}
    \end{align*}
    This way, $[[Γ ⊢ [σ1]β± ≈ β±]]$ for any $[[β±]] \in [[fv ∀pas.iM]]$ and thus,
    $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$.
